---
title: "Feature Engineering for EEG Seizure Detection\n"
author: "22110012 - Huu An  \n\n 22110059 - Hong Hien\n"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: false
    theme: united
    highlight: tango
    mathjax: default
  pdata_document:
    toc: true
    toc_depth: '3'
---

```{css, echo=FALSE}
.tocify {
  background-color: #f0f8ff;  /* Sidebar background color (Alice Blue) */
  border-radius: 8px;
  padding: 10px;
  border: 1px solid #cce;
  box-shadow: 2px 2px 5px rgba(0,0,0,0.1);
}

.tocify-header {
  color: #004085; /* TOC header color */
  font-weight: bold;
}

.tocify-item.active {
  background-color: #d1ecf1; /* Active item background color */
}
```

```{r setup, include=FALSE}
# --- Package installation list ---
# install.packages(c(
#   "tidyverse", "moments", "seewave", "wavelets", "TSEntropies",
#   "caret", "themis", "recipes", "randomForest", "e1071",
#   "patchwork", "pracma", "pROC", "Rtsne", "MASS",
#   "psych", "uwot", "biotools", "mvnormtest"
# ))

# --- Global chunk settings ---
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center')

# --- Data manipulation and visualization ---
library(tidyverse)    # Data wrangling, ggplot2
library(patchwork)    # Combining ggplot figures

# --- Statistical analysis and hypothesis testing ---
library(moments)      # Skewness, kurtosis
library(psych)        # KMO, Bartlett tests, FA
library(biotools)     # Box's M test
library(mvnormtest)   # Multivariate normality tests

# --- Signal processing and feature extraction ---
library(seewave)      # Signal and spectral analysis
library(wavelets)     # Discrete Wavelet Transform
library(TSEntropies)  # Time series entropy calculation
library(pracma)       # Trapezoidal integration (trapz), mathematical utilities

# --- Machine learning and preprocessing ---
library(caret)        # ML framework
library(themis)       # SMOTE for class imbalance handling
library(recipes)      # Preprocessing pipelines
library(randomForest) # Random Forest model
library(e1071)        # SVM model

# --- Dimensionality reduction and manifold visualization ---
library(Rtsne)        # t-SNE
library(uwot)         # UMAP
library(MASS)         # LDA
```

# I. Project Overview

**1.1. Background and Research Problem**

Electroencephalography (EEG) is an indispensable clinical tool in the diagnosis and monitoring of epilepsy—a neurological disorder characterized by recurrent, unprovoked seizures resulting from excessive, synchronous electrical discharges in cerebral neurons. Timely and accurate detection of epileptic seizures is paramount for effective treatment planning, medication management, and improving patients' overall quality of life.

However, traditional visual inspection of EEG recordings presents significant operational bottlenecks: it is labor-intensive, time-consuming, highly reliant on clinician expertise (introducing subjective bias), and inherently difficult to scale for continuous long-term monitoring. These challenges necessitate reliable automated computational solutions.

To address these limitations, this study designs, implements, and evaluates an automated machine learning framework capable of capturing salient latent patterns from short-term EEG segments to accurately classify seizure vs. non-seizure activity.

**1.2. Research Questions and Objectives**

This research is guided by two fundamental questions:

1. Can features systematically extracted from the time, frequency, and wavelet domains reliably discriminate between seizure and non-seizure states from short-term (1-second) EEG epochs?

2. How does dimensionality reduction, specifically Principal Component Analysis (PCA), impact the trade-off between classification accuracy and computational complexity?

Guided by these questions, the ultimate objective of this study is to construct, benchmark, and validate the most effective classification architecture, providing a rigorous methodological framework for automated seizure detection.

---

# II. Data and Methodology

**2.1. Dataset Description**

**Data Origin**

This study employs the publicly available **"Epileptic Seizure Recognition"** dataset hosted on Kaggle. This dataset is a preprocessed and restructured version of the renowned benchmark database collected by Andrzejak et al. (2001) at the University of Bonn.

**Original Recording Structure**

The original database consists of five distinct sets (denoted A through E), each containing 100 single-channel EEG segments recorded across five different clinical conditions:

* **Set E (Corresponding to original label y = 1):** Recorded intracranially **during epileptic seizure episodes** (ictal state).

* **Set D (Corresponding to original label y = 2):** Intracranial recordings from the epileptogenic zone of epilepsy patients during seizure-free intervals (interictal state).

* **Set C (Corresponding to original label y = 3):** Intracranial recordings from the hippocampal formation of the opposite (healthy) hemisphere during seizure-free intervals (interictal state).

* **Set B (Corresponding to original label y = 4):** Surface EEG recordings from healthy volunteers with **eyes closed**.

* **Set A (Corresponding to original label y = 5):** Surface EEG recordings from healthy volunteers with **eyes open**.

Each original recording was sampled at 173.61 Hz across 4,097 data points (~23.6 seconds duration).

**Kaggle Dataset Construction**

[Kaggle Epileptic Seizure Recognition Dataset](https://www.kaggle.com/datasets/yasserhessein/epileptic-seizure-recognition/data)

To generate the `.csv` dataset used in this study, the following processing steps were applied:

1. Each continuous recording (4,097 points) was divided into 23 non-overlapping epochs.

2. Each epoch contains **178 data points**, corresponding to approximately **1 second** of EEG activity.

3. This process generated a total of **11,500 samples** (500 recordings × 23 epochs/recording), each forming one row in the dataset.

Thus, the final dataset contains:

* **Dimensions:** 11,500 rows (samples), where each row is a 1-second EEG segment.

* **Predictors:** 178 numerical feature columns (`X1` through `X178`), representing voltage amplitudes.

* **Target Label:** A categorical variable `y` ranging from 1 to 5, corresponding to the original five recording groups.

**2.2. Formulation of the Binary Classification Problem**

To align with the clinical objective of rapid seizure detection, the multi-class problem is cast as a binary classification task:

* **Class 1 (Seizure):** EEG epochs recorded during an active epileptic seizure (original label `y = 1`).
* **Class 0 (Non-Seizure):** All other EEG epochs, including normal brain states with eyes open/closed and interictal recordings (original labels `y = 2, 3, 4, 5`).

**2.3. Overall Research Pipeline**

The investigation follows a 4-phase methodology:

1. **Data Preparation:** Binarize target labels and partition data into training (80%) and test (20%) sets.

2. **Feature Engineering:** Extract 39 discriminative features across the time domain, frequency domain (power spectral density across delta, theta, alpha, beta, and gamma bands), and wavelet domain (Daubechies-4 discrete wavelet decomposition).

3. **Exploratory Data Analysis and Dimensionality Reduction:** Apply multivariate statistics (KMO, Bartlett, Factor Analysis, MANOVA) and non-linear visualization techniques (t-SNE, UMAP) to evaluate feature separability, alongside PCA scree plot analysis.

4. **Model Training and Evaluation:** Mitigate class imbalance using SMOTE on the training set, followed by 10-fold cross-validation across LDA, SVM (Radial), and Random Forest. Evaluate top candidate models on the held-out test set.

```{r load-data}
# Load dataset
df <- readr::read_csv("Epileptic Seizure Recognition.csv")

# Remove the initial identifier column
df <- df %>% dplyr::select(-1)

# Binarize the target variable (1 for seizure, 0 for non-seizure)
df <- df %>%
  mutate(y = ifelse(y == 1, 1, 0))

# Display the first 6 rows
head(df)

# Separate predictors (X) and target variable (y)
X_original <- df %>% dplyr::select(-y)
y_labels <- df$y
```

```{r}
library(patchwork) # For combining ggplot figures

# Extract a random Non-Seizure sample (y = 0)
non_seizure_sample <- df %>% filter(y == 0) %>% sample_n(1) %>% dplyr::select(-y) %>% as.numeric()
# Extract a random Seizure sample (y = 1)
seizure_sample <- df %>% filter(y == 1) %>% sample_n(1) %>% dplyr::select(-y) %>% as.numeric()

# Helper plotting function
plot_eeg <- function(signal, plot_title) {
  tibble(time_step = 1:length(signal), amplitude = signal) %>%
    ggplot(aes(x = time_step, y = amplitude)) +
    geom_line() +
    labs(title = plot_title, x = "Time Step", y = "Amplitude (µV)") +
    theme_minimal()
}

# Generate both plots
p1 <- plot_eeg(non_seizure_sample, "Sample Signal: Non-Seizure")
p2 <- plot_eeg(seizure_sample, "Sample Signal: Seizure")

# Display side-by-side comparison
p1 + p2
```

# III. Feature Engineering

## 3.1. Time Domain Features

Prior to extraction, we visualize a random signal epoch in the time domain:
```{r plot-signal}
# Extract a random sample from X
set.seed(42) # For reproducibility
signal_sample <- X_original %>% dplyr::sample_n(1) %>% as.numeric()

# Create time domain waveform plot
tibble(
  time_step = 1:length(signal_sample),
  amplitude = signal_sample
) %>%
  ggplot(aes(x = time_step, y = amplitude)) +
  geom_line(color = "dodgerblue") +
  geom_point(color = "dodgerblue", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", alpha = 0.5) +
  labs(
    title = "Sample 1-Second EEG Signal Waveform",
    x = "Time Step",
    y = "Amplitude (µV)"
  ) +
  theme_minimal() +
  theme(panel.grid.major = element_line(linetype = "dashed", color = "gray80"),
        panel.grid.minor = element_blank())
```

This visualization displays the `time domain` signal, capturing instantaneous amplitude variations over time. From this representation, we extract standard and advanced statistical metrics that quantify signal shape, amplitude dispersion, and morphology.

```{r time-features-function}
# Function to extract time domain features
extract_time_domain_features <- function(signal) {
  # Basic statistical metrics
  mean_val <- mean(signal)
  std_val <- sd(signal)
  var_val <- var(signal)
  min_val <- min(signal)
  max_val <- max(signal)
  skew_val <- moments::skewness(signal)
  kurt_val <- moments::kurtosis(signal)
  rms_val <- sqrt(mean(signal^2))
  zero_crossings <- sum(diff(sign(signal)) != 0)
  
  # Maximum absolute amplitude
  abs_max <- max(abs(c(min_val, max_val)))
  
  # Waveform shape factors
  crest_factor <- abs_max / rms_val
  margin_factor <- abs_max / var_val
  shape_factor <- rms_val / mean(abs(signal))
  impulse_factor <- abs_max / mean(abs(signal))
  
  # Return as a tibble
  tibble(
    mean = mean_val, std = std_val, var = var_val, min = min_val, max = max_val,
    skew = skew_val, kurtosis = kurt_val, rms = rms_val,
    zero_crossings = zero_crossings, abs_max = abs_max,
    crest_factor = crest_factor, margin_factor = margin_factor,
    shape_factor = shape_factor, impulse_factor = impulse_factor
  )
}
```

**Crest factor** is defined as the ratio between the maximum absolute amplitude and the root mean square (RMS) value:

$$
C(y_n) = \frac{\max |y_n|}{RMS(y_n)} \qquad \text{where } RMS(y_n) = \sqrt{\frac{1}{N} \sum_{n=1}^N y_n^2}
$$

This metric quantifies how extreme the signal peaks are relative to its average power.

**Margin factor** similarly compares peak amplitude against signal variance:

$$
M = \frac{\max |y_n|}{\text{Var}(y_n)}, \qquad \text{where } \text{Var}(y_n) = \frac{1}{N} \sum_{n=1}^N \left( y_n - \bar{y} \right)^2
$$

It reflects the magnitude of the highest peak in relation to signal dispersion. When the signal mean is zero (as in zero-centered Gaussian noise), crest and margin factors exhibit strong correlation.

**Shape factor** measures the ratio of the RMS value to the Mean Absolute Value (MAV):

$$
S(y_n) = \frac{\text{RMS}}{\text{MAV}}, \quad \text{where } \text{MAV} = \frac{1}{N} \sum_{n=1}^N |y_n|
$$

**Impulse factor** compares the maximum absolute peak amplitude to the MAV:

$$
I = \frac{\max |y_n|}{\text{MAV}}
$$

## 3.2. Frequency Domain Features

Transforming EEG signals into the frequency domain allows the analysis of spectral energy distribution across standard physiological frequency bands. Frequency analysis reveals underlying rhythmic oscillations, cognitive states, and abnormal paroxysmal discharges associated with epilepsy.

Welch's method estimates the **Power Spectral Density (PSD)** by averaging modified periodograms over overlapping windowed segments, reducing spectral variance:

$$
S_{xx}(f) = \frac{1}{K} \sum_{k=1}^{K} \left| X_k(f) \right|^2
$$

where $K$ denotes the number of segments and $X_k(f)$ is the Discrete Fourier Transform of the windowed $k$-th segment.

| Spectral Band | Frequency Range (Hz) | Clinical & Physiological Description |
|:--------------|:--------------------:|:-------------------------------------|
| **Delta (δ)** | 0.5 – 4 Hz | Associated with deep sleep and regenerative states. Abnormally elevated awake delta may reflect brain lesions or epileptic slowing. |
| **Theta (θ)** | 4 – 8 Hz | Correlated with drowsiness, light sleep, and memory consolidation. Elevated theta often precedes or accompanies epileptiform activity. |
| **Alpha (α)** | 8 – 13 Hz | Predominant during relaxed wakefulness with eyes closed. Desynchronization occurs during focal seizures and cognitive tasks. |
| **Beta (β)** | 13 – 30 Hz | Linked with active alertness, problem-solving, and focus. Fast rhythmic bursts may occur during focal cortical seizures. |
| **Gamma (γ)** | 30 – 89 Hz | Associated with complex sensory binding and high-level cognitive processing. High-frequency oscillations often characterize ictal onset. |

```{r freq-features-function}
# Function to extract frequency domain band powers
extract_frequency_domain_features <- function(signal, fs = 178) {
  # Compute mean spectrum via seewave::meanspec
  mean_spec_res <- seewave::meanspec(signal, f = fs, wl = length(signal), plot = FALSE)
  
  freqs <- mean_spec_res[, 1] * 1000 # Convert from kHz to Hz
  psd <- mean_spec_res[, 2]
  
  # Helper function to compute band power using trapezoidal integration
  bandpower <- function(psd, freqs, fmin, fmax) {
    idx <- freqs >= fmin & freqs <= fmax
    pracma::trapz(freqs[idx], psd[idx])
  }
  
  # Compute standard EEG frequency band powers (Hz)
  delta_power <- bandpower(psd, freqs, 0.5, 4)
  theta_power <- bandpower(psd, freqs, 4, 8)
  alpha_power <- bandpower(psd, freqs, 8, 13)
  beta_power  <- bandpower(psd, freqs, 13, 30)
  gamma_power <- bandpower(psd, freqs, 30, fs / 2)
  
  tibble(
    delta_power = delta_power, theta_power = theta_power,
    alpha_power = alpha_power, beta_power = beta_power,
    gamma_power = gamma_power
  )
}
```

```{r}
# Sample signal spectral visualization
set.seed(42)
signal_sample <- X_original %>% dplyr::sample_n(1) %>% as.numeric()
fs <- 178 # Sampling frequency (Hz)

# Calculate PSD
mean_spec_res <- seewave::meanspec(signal_sample, f = fs, wl = length(signal_sample), plot = FALSE)

spectrum_df <- tibble(
  frequency = mean_spec_res[, 1] * 1000, # kHz to Hz
  amplitude = mean_spec_res[, 2]
)

# Band boundary specifications
eeg_bands <- tibble(
  band = factor(c("Delta", "Theta", "Alpha", "Beta", "Gamma"), levels = c("Delta", "Theta", "Alpha", "Beta", "Gamma")),
  fmin = c(0.5, 4, 8, 13, 30),
  fmax = c(4, 8, 13, 30, fs / 2)
)

# Render spectral plot
ggplot(spectrum_df, aes(x = frequency, y = amplitude)) +
  geom_rect(data = eeg_bands, aes(xmin = fmin, xmax = fmax, ymin = -Inf, ymax = Inf, fill = band), 
            alpha = 0.3, inherit.aes = FALSE) +
  geom_line(color = "black") +
  coord_cartesian(xlim = c(0, 50)) +
  scale_fill_brewer(palette = "Set1", name = "EEG Band") +
  labs(
    title = "Power Spectral Density of Sample Signal",
    subtitle = "Energy distribution across physiological EEG frequency bands",
    x = "Frequency (Hz)",
    y = "Spectral Amplitude"
  ) +
  theme_minimal()
```

## 3.3. Wavelet and Non-linear Features

The Wavelet Transform (WT) provides joint time-frequency localization, making it exceptionally well-suited for non-stationary bio-signals such as EEG. Discrete Wavelet Transform (DWT) decomposes the signal into hierarchical approximation and detail coefficients:
- **Low-frequency components (approximation coefficients)** capture baseline trends and slow rhythmic activity.
- **High-frequency components (detail coefficients)** detect transient spikes, sharp waves, and localized paroxysmal discharges.

Here, we decompose signals using a Daubechies-4 (`db4`) wavelet across 4 decomposition levels, extracting coefficient energy, standard deviation, skewness, and kurtosis.

Additionally, **Approximate Entropy (ApEn)** quantifies regularity, predictability, and algorithmic complexity in the time series:

$$
ApEn(m, r) = \phi(m) - \phi(m+1)
$$

where $m$ denotes embedding dimension and $r$ is the filtering tolerance threshold. A higher ApEn indicates higher complexity/irregularity, whereas lower values reflect predictable, synchronized oscillations (characteristic of certain seizure patterns).

```{r wavelet-features-function}
extract_wavelet_nonlinear_features <- function(signal) {

  # Ensure signal length is a power of 2 for DWT
  if (log2(length(signal)) %% 1 != 0) {
      new_len <- 2^floor(log2(length(signal)))
      signal_for_dwt <- signal[1:new_len]
  } else {
      signal_for_dwt <- signal
  }
  
  dwt_res <- wavelets::dwt(signal_for_dwt, filter = "d4", n.levels = 4)
  wavelet_coeffs <- c(dwt_res@W, list(dwt_res@V[[4]]))
  level_names <- paste0("level_", 1:length(wavelet_coeffs))
  
  all_features <- list()

  # Iterate across levels to extract energy and statistical metrics
  for (i in 1:length(wavelet_coeffs)) {
    coeffs <- wavelet_coeffs[[i]]
    level_name <- level_names[i]
    
    all_features[[paste0(level_name, "_energy")]] <- sum(coeffs^2)
    all_features[[paste0(level_name, "_std")]]    <- sd(coeffs)
    all_features[[paste0(level_name, "_skew")]]   <- moments::skewness(coeffs)
    all_features[[paste0(level_name, "_kurt")]]   <- moments::kurtosis(coeffs)
  }
  
  return(as_tibble(all_features))
}
```

```{r}
# Extract wavelet features from sample signal
wavelet_features <- extract_wavelet_nonlinear_features(signal_sample)

# Transform into long format for visualization
wavelet_energy_df <- wavelet_features %>%
  dplyr::select(ends_with("_energy")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "level",
    values_to = "energy"
  ) %>%
  mutate(level = sub("_energy", "", level) %>% sub("_", " ", .) %>% str_to_title())

# Render bar chart
ggplot(wavelet_energy_df, aes(x = level, y = energy, fill = level)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = round(energy, 2)), vjust = -0.5) +
  labs(
    title = "Signal Energy Distribution Across Wavelet Decomposition Levels",
    x = "Wavelet Decomposition Level",
    y = "Energy (Sum of Squared Coefficients)"
  ) +
  theme_minimal()
```

---

## 3.4. Feature Computation and Dataset Construction

```{r apply-feature-engineering-and-clean}
# Apply feature extraction across all samples
features_df_time <- purrr::map_dfr(1:nrow(X_original), ~extract_time_domain_features(as.numeric(X_original[.x, ])))
features_df_freq <- purrr::map_dfr(1:nrow(X_original), ~extract_frequency_domain_features(as.numeric(X_original[.x, ])))
features_df_wavelet <- purrr::map_dfr(1:nrow(X_original), ~extract_wavelet_nonlinear_features(as.numeric(X_original[.x, ])))

# Combine extracted feature spaces
features_df <- bind_cols(features_df_time, features_df_freq, features_df_wavelet)

# Display preview of extracted features
knitr::kable(head(features_df))
```