---
title: "Exploratory Data Analysis and Dimensionality Reduction\n"
author: "22110012 - Huu An  \n\n 22110059 - Hong Hien\n"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: false
    theme: united
    highlight: tango
    mathjax: default
  pdata_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include=FALSE}
# --- Global chunk settings ---
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center')

# --- Libraries for multivariate analysis and dimensionality reduction ---
library(tidyverse)    # Data manipulation and ggplot2
library(caret)        # Preprocessing (nearZeroVar, preProcess)
library(factoextra)   # PCA visualization
library(psych)        # KMO, Bartlett tests, Factor Analysis (FA)
library(biotools)     # Box's M test
library(MVN)          # Multivariate normality testing
library(Rtsne)        # Non-linear dimensionality reduction: t-SNE
library(uwot)         # Non-linear dimensionality reduction: UMAP
library(patchwork)    # Combining plots
library(scales)       # Percentage scaling formatting
```

# IV. Multivariate Statistical Analysis

We explore the newly engineered feature representations using multivariate statistical methods.

```{r}
# --- IV. Multivariate Statistical Analysis ---

# Load extracted feature dataset
train_data <- readr::read_csv("EEG_Features_Extracted.csv")

# Separate predictors (X) and target variable (y)
X_features <- train_data %>% dplyr::select(-target)
y_labels_factor <- as.factor(train_data$target)

# Verify data dimensions
# Expected: 11,500 rows and 39 feature columns
dim(X_features)
```

```{r prepare-for-analysis}
# Identify near-zero variance predictors
nzv_cols_indices <- nearZeroVar(X_features)

# Report removed columns (if any)
if(length(nzv_cols_indices) > 0) {
  print("Removing near-zero variance columns:")
  print(names(X_features)[nzv_cols_indices])
  
  # Exclude these columns from analysis
  X_features_filtered <- X_features[, -nzv_cols_indices]
} else {
  X_features_filtered <- X_features
}

# Standardize (center and scale) the filtered feature space
scaler <- preProcess(X_features_filtered, method = c("center", "scale"))
X_features_scaled <- predict(scaler, X_features_filtered)

# Check for non-finite values
print(paste("Remaining non-finite values:", sum(!is.finite(as.matrix(X_features_scaled)))))
```

## 4.1. Principal Component Analysis (PCA)

PCA reduces feature dimensionality while preserving maximal variance and eliminating multicollinearity.

```{r pca-analysis}
pca_obj <- prcomp(X_features_scaled, scale. = TRUE)
```

```{r plot}
# Extract explained and cumulative variance
explained_variance <- pca_obj$sdev^2 / sum(pca_obj$sdev^2)
cumulative_variance <- cumsum(explained_variance)

# Construct dataframe for visualization
plot_data <- tibble(
  component = 1:length(explained_variance),
  explained_variance = explained_variance,
  cumulative_variance = cumulative_variance
)

# Combined Scree and Cumulative Variance Plot
ggplot(plot_data, aes(x = component)) +
  # Individual variance bars
  geom_col(aes(y = explained_variance), fill = "steelblue", alpha = 0.8) +
  
  # Elbow line
  geom_line(aes(y = explained_variance), color = "black", size = 0.8) +
  geom_point(aes(y = explained_variance), color = "black", size = 2) +
  
  # Cumulative variance curve
  geom_line(aes(y = cumulative_variance), color = "red", size = 1) +
  geom_point(aes(y = cumulative_variance), color = "red", size = 2) +
  
  # Percentage labels for first 10 components
  geom_text(
    data = . %>% filter(component <= 10),
    aes(y = explained_variance, label = scales::percent(explained_variance, accuracy = 0.1)),
    vjust = -0.8,
    size = 3.5,
    color = "black"
  ) +
  
  # Reference thresholds
  geom_hline(yintercept = 0.9, linetype = "dashed", color = "blue", alpha = 0.7) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "darkgreen", alpha = 0.7) +
  annotate("text", x = max(plot_data$component) * 0.7, y = 0.9, label = "90% Variance Threshold", vjust = -0.5, color = "blue") +
  annotate("text", x = max(plot_data$component) * 0.7, y = 0.95, label = "95% Variance Threshold", vjust = -0.5, color = "darkgreen") +
  
  scale_x_continuous(breaks = 1:length(explained_variance)) +
  scale_y_continuous(
    name = "Proportion of Explained Variance",
    labels = scales::percent
  ) +
  labs(
    title = "Combined PCA Scree Plot",
    subtitle = "Individual variance (bars, black line) and cumulative variance (red curve)",
    x = "Principal Component",
    y = "Proportion of Explained Variance"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )
```

**Interpretation:**

The Scree Plot offers two standard approaches for selecting the number of principal components:

1. **Elbow Criterion:**
   The variance reduction flattens noticeably at PC3 or PC4, indicating that subsequent components capture progressively diminishing amounts of unique variance.

2. **Cumulative Variance Threshold:**
   - Retaining **90%** of total variance requires **9 PCs**.
   - Retaining **95%** of total variance requires **12 PCs**.

**Conclusion:**

* **Data Interpretation:** The Elbow Method suggests **3–4 PCs** provide a parsimonious summary for visual exploration and latent structure interpretation.
* **Model Training:** Retaining **9–12 PCs** preserves **90–95%** of cumulative variance, providing sufficient discriminative information for machine learning while substantially reducing computational overhead and noise.

```{r}
library(factoextra)

# PCA Scatter Plot (PC1 vs PC2)
fviz_pca_ind(pca_obj,
             geom.ind = "point",
             col.ind = y_labels_factor,
             palette = c("#00AFBB", "#E7B800"),
             addEllipses = TRUE,
             legend.title = "Class") +
  labs(title = "PCA Observation Projection (PC1 vs PC2)")
```

*Observation:*
The scatter plot demonstrates noticeable separation between Class 0 (Non-Seizure) and Class 1 (Seizure) across the first two principal components, which account for 56.7% of total variance (39.7% from PC1 and 17.0% from PC2). Separation occurs predominantly along the first dimension (Dim1). Some overlap is visible near the decision boundary, suggesting that non-linear manifold representations (e.g., t-SNE, UMAP) are warranted to inspect non-linear separability.

```{r}
eigenvalues <- pca_obj$sdev^2
variance_percent <- eigenvalues / sum(eigenvalues) * 100

data.frame(
  PC = paste0("PC", 1:length(eigenvalues)),
  Eigenvalue = round(eigenvalues, 2),
  Variance_Percent = round(variance_percent, 2),
  Cumulative_Percent = round(cumsum(variance_percent), 2)
)
```

Applying the Kaiser-Guttman criterion (eigenvalues > 1):
```{r}
eigenvalues <- pca_obj$sdev^2
selected_pcs <- sum(eigenvalues > 1)
selected_pcs
```

```{r}
# Extract loadings for the first 4 PCs in tidy format
loadings_long <- as.data.frame(pca_obj$rotation[, 1:4]) %>%
  mutate(feature = rownames(.)) %>%
  tidyr::pivot_longer(
    cols = c(PC1, PC2, PC3, PC4),
    names_to = "pc",
    values_to = "loading"
  )

# Render separate loadings bar charts for each component
pc_names <- unique(loadings_long$pc)

for (pc_name in pc_names) {
  plot_data <- loadings_long %>% filter(pc == pc_name)
  
  p <- ggplot(plot_data, aes(y = loading, 
                             x = reorder(feature, loading), 
                             fill = loading > 0)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    labs(
      title = paste("Factor Loading Analysis for", pc_name),
      x = "Feature",
      y = "Loading Value"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 16, face = "bold")
    )
  
  print(p)
}
```

**Interpretation of Principal Components:**

**PC1 - Overall Signal Magnitude and Energy**
PC1 captures the primary amplitude and energy dispersion of the EEG recording. It correlates strongly with amplitude scale features: `abs_max`, `std`, `rms`, and `var`. All wavelet decomposition energies (`wavelet_energy_level_1` through `5`) display high positive loadings, highlighting PC1's role in measuring broad-spectrum energy. It effectively differentiates high-energy, volatile seizure activity from low-amplitude baseline recordings.
* **Strong Positive Loadings:** `abs_max`, `std`, `rms`, `var`, all `wavelet_energy_level` features.
* **Strong Negative Loadings:** `min`, `skew`.

---

**PC2 - Contrast Between Raw Power and Spectral/Waveform Morphology**
PC2 captures a contrast between raw amplitude power and peak morphology. It exhibits positive loadings on bulk energy indicators (`rms`, `var`, `wavelet_energy_level_4`) but negative loadings on peakedness factors (`crest_factor`, `margin_factor`) and discrete band powers (`delta`, `theta`, `alpha`, `beta`, `gamma`).
* **Strong Positive Loadings:** `wavelet_energy_level_4`, `rms`, `var`.
* **Strong Negative Loadings:** `crest_factor`, `impulse_factor`, all discrete spectral band powers.

---

**PC3 - Signal Complexity and High-Frequency Rhythms**
PC3 highlights cognitive and neurophysiological complexity. It exhibits strong positive correlation with dynamical complexity indicators: `approx_entropy`, `zero_crossings`, and fast frequency band powers (`gamma_power`, `beta_power`, `alpha_power`). Conversely, it loads negatively on `kurtosis` and impulsive peak ratios.
* **Strong Positive Loadings:** `approx_entropy`, `gamma_power`, `beta_power`, `zero_crossings`.
* **Strong Negative Loadings:** `kurtosis`, `crest_factor`, `impulse_factor`.

---

**PC4 - Transient Spikiness and Low-to-Mid Frequency Characteristics**
PC4 reflects impulsive wave morphology and transient sharpness versus smooth low-frequency rhythmic oscillations. It loads positively on `margin_factor` and `kurtosis`, and negatively on lower frequency bands (`delta_power`, `theta_power`, `alpha_power`) and `skew`.
* **Strong Positive Loadings:** `margin_factor`, `kurtosis`, `approx_entropy`.
* **Strong Negative Loadings:** `skew`, `delta_power`, `theta_power`, `alpha_power`.

---

## 4.2. Factor Analysis (FA)

Factor Analysis identifies latent unobservable constructs underlying the observed feature correlations.

```{r factor-analysis}
# KMO Test for sampling adequacy
cat("--- KMO Test Results ---\n")
kmo_result <- KMO(X_features_scaled)
print(kmo_result)

# Bartlett's Test of Sphericity
cat("\n--- Bartlett's Test Results ---\n")
bartlett_result <- cortest.bartlett(X_features_scaled)
print(bartlett_result)
```

*Observation:*
Both diagnostic tests confirm data adequacy for Factor Analysis. The overall Kaiser-Meyer-Olkin Measure of Sampling Adequacy (Overall MSA) is **0.82** ("meritorious"), indicating strong common variance among predictors. Bartlett's test of sphericity is highly significant ($p < 2.2 \times 10^{-16}$), rejecting the identity correlation matrix hypothesis.

```{r Parallel Analysis}
library(psych)

# Parallel Analysis
fa.parallel(X_features_scaled, fa = "fa")

# Velicer's MAP
VSS(X_features_scaled)
```

The Parallel Analysis scree plot compares empirical eigenvalues from observed data against eigenvalues from random simulated matrices. The empirical eigenvalues substantially exceed simulated thresholds for the **first 7 factors**, confirming that retaining **7 latent factors** is statistically optimal.

```{r}
# Execute Factor Analysis with 7 factors and Varimax rotation
fa_result <- fa(X_features_scaled, nfactors = 7, rotate = "varimax")

# Display factor loadings matrix (suppressing loadings < 0.4 for clarity)
cat("\n--- Factor Analysis Loadings Matrix ---\n")
print(fa_result$loadings, cutoff = 0.4)
```

**Interpretation of Latent Factors:**

* **Factor 1 (MR1) - Overall Energy and Amplitude (35.8% Variance):** Dominated by `std` (0.975), `var` (0.985), `min` (-0.920), `max` (0.879), `rms` (0.984), `abs_max` (0.944), and wavelet energies. Represents total signal power and volatility.
* **Factor 2 (MR2) - Peak Sharpness and Waveform Morphology (13.2% Variance):** Characterized by `kurtosis` (0.814), `crest_factor` (0.905), `shape_factor` (0.760), and `impulse_factor` (0.973). Captures paroxysmal spike transients.
* **Factor 3 (MR3) - High-Frequency Rhythms and Oscillation Rate (12.5% Variance):** Governed by `zero_crossings` (0.852), `alpha_power` (0.716), `beta_power` (0.830), and `gamma_power` (0.757). Represents fast cerebral rhythms.
* **Factor 4 (MR4) - Low-Frequency Rhythms:** Associated with `margin_factor` (0.613).
* **Factor 5 (MR5) - Delta Band Power:** Dominated by `delta_power` (0.885), capturing slow-wave delta synchronization.
* **Factor 6 (MR6) - Signal Skewness:** Characterized by distribution asymmetry (`skew`: 0.731).
* **Factor 7 (MR7) - High-Resolution Wavelet Residuals:** Captures localized high-frequency detail coefficients (`wavelet_energy_level_1`, `level_2`).

*Cumulative Variance:* The 7 latent factors account for **78.0% of total cumulative variance**, condensing the feature space into interpretable neurophysiological dimensions.

```{r}
library(tidytext)

# Extract and pivot factor loadings
loadings_matrix <- fa_result$loadings
loadings_long <- as.data.frame(loadings_matrix[,]) %>%
  mutate(feature = rownames(.)) %>%
  pivot_longer(
    cols = -feature,
    names_to = "factor",
    values_to = "loading"
  )

# Plot factor loadings across each latent construct
factor_names <- unique(loadings_long$factor)

for (factor_name in factor_names) {
  plot_data <- loadings_long %>% filter(factor == factor_name)
  
  p <- ggplot(plot_data, aes(y = loading, 
                             x = reorder(feature, loading), 
                             fill = loading > 0)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    labs(
      title = paste("Feature Loadings for Factor", factor_name),
      x = "Feature",
      y = "Loading Value"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 16, face = "bold")
    )
  
  print(p)
}
```

---

```{r}
# Extract factor scores
factor_scores <- as.data.frame(fa_result$scores)

# Merge factor scores with class labels
fa_data_for_plot <- bind_cols(factor_scores, Class = y_labels_factor)

# Scatter plot of first two factors (MR1 vs MR2)
ggplot(fa_data_for_plot, aes(x = MR1, y = MR2, color = Class)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Factor Analysis Latent Score Projection",
    subtitle = "Sample distribution across MR1 (Energy/Amplitude) and MR2 (Peak Morphology)",
    x = "Factor 1 (MR1): Overall Energy & Amplitude",
    y = "Factor 2 (MR2): Peak Characteristics & Waveform Shape"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("0" = "#00AFBB", "1" = "#E7B800"), name = "Class") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5)
```

**Observation:**
While distinct clusters emerge along MR1 and MR2, significant overlap remains in the linear two-factor projection, reinforcing the need for non-linear manifold exploration.

## 4.3. Multivariate Analysis of Variance (MANOVA)

MANOVA formally tests whether group mean vectors across all features differ significantly between the Seizure and Non-Seizure cohorts.

### 4.3.1. Diagnostic Assumption Testing

1. *Homogeneity of Covariance Matrices (Box's M Test)*
- $H_0$: Covariance matrices are equal across groups ($\Sigma_1 = \Sigma_0$).
- $H_a$: Covariance matrices differ ($\Sigma_1 \ne \Sigma_0$).

```{r}
boxm_result <- boxM(data = X_features_scaled, grouping = y_labels_factor)
print(boxm_result)
```

**Observation:** Box’s M yields $p < 2.2 \times 10^{-16}$, rejecting $H_0$. Because Box's M is exceptionally sensitive to large sample sizes ($N = 11,500$), we rely on **Pillai’s Trace**, which is robust to covariance heterogeneity.

2. *Multivariate Normality Testing (Henze-Zirkler)*
- $H_0$: Data follows a multivariate normal distribution.
- $H_a$: Data deviates from multivariate normality.

```{r}
library(MVN)

cat("--- Multivariate Normality Test: Non-Seizure Group (0) ---\n")
MVN::mvn(data = X_features_scaled[y_labels_factor == "0", ])

cat("\n--- Multivariate Normality Test: Seizure Group (1) ---\n")
MVN::mvn(data = X_features_scaled[y_labels_factor == "1", ])
```

**Observation:** The Henze-Zirkler test rejects multivariate normality for both classes ($p < 0.001$).

### 4.3.2. Executing MANOVA Test

Hypotheses:
* **$H_0$**: Population mean vectors of features are identical across classes: $\boldsymbol{\mu}_{\text{seizure}} = \boldsymbol{\mu}_{\text{non-seizure}}$
* **$H_a$**: At least one feature mean differs: $\boldsymbol{\mu}_{\text{seizure}} \ne \boldsymbol{\mu}_{\text{non-seizure}}$

```{r manova-analysis}
# Execute MANOVA
manova_result <- manova(as.matrix(X_features_scaled) ~ y_labels_factor)
summary(manova_result, test = "Pillai")
```

**Conclusion:** Pillai's Trace statistic is highly significant ($p < 2.2 \times 10^{-16}$), confirming that feature centroids between seizure and non-seizure populations differ substantially.

# V. Non-linear Manifold Visualizations

Given the non-linear boundaries identified in linear projections (PCA/FA), we employ **t-SNE** and **UMAP** to project high-dimensional feature spaces onto 2D while preserving localized neighborhood topology.

## 5.1. t-SNE and UMAP Projections

```{r tsne-umap-analysis}
set.seed(42) # For reproducibility

# t-SNE on full dataset
cat("Running t-SNE on 11,500 samples...\n")
tsne_out <- Rtsne(X_features_scaled, dims = 2, perplexity = 30, verbose = FALSE, max_iter = 500)
df_tsne <- tibble(Dim1 = tsne_out$Y[,1], Dim2 = tsne_out$Y[,2], Class = y_labels_factor)
p_tsne <- ggplot(df_tsne, aes(x = Dim1, y = Dim2, color = Class)) + 
  geom_point(alpha = 0.7) + 
  labs(title = "t-SNE Projection (Full Dataset)") + 
  theme_minimal()

# UMAP on full dataset
cat("Running UMAP on 11,500 samples...\n")
umap_out <- uwot::umap(X_features_scaled, n_neighbors = 15, min_dist = 0.1, n_components = 2)
df_umap <- tibble(Dim1 = umap_out[,1], Dim2 = umap_out[,2], Class = y_labels_factor)
p_umap <- ggplot(df_umap, aes(x = Dim1, y = Dim2, color = Class)) + 
  geom_point(alpha = 0.7) + 
  labs(title = "UMAP Projection (Full Dataset)") + 
  theme_minimal()

# Side-by-side comparison
cat("Rendering manifold plots...\n")
p_tsne + p_umap
```

**Observation:** Both t-SNE and UMAP reveal stark cluster separation between Seizure and Non-Seizure cohorts with minimal inter-class overlap. UMAP in particular preserves continuous manifold trajectories. This demonstrates that the engineered feature set possesses strong non-linear discriminatory capacity.

## 5.2. Comparative Synthesis: PCA, FA, t-SNE, and UMAP

* **PCA:** Preserves global variance (56.7% in 2 components), separating classes along Dim1, but constrained by linear orthogonality.
* **Factor Analysis:** Illuminates 7 neurophysiological latent dimensions (78.0% cumulative variance).
* **t-SNE and UMAP:** Confirm clear, non-linear class separation in low dimensions.
* **Implications for Modeling:** The non-linear separability suggests kernelized non-linear classifiers (such as SVM-RBF or Random Forest) will achieve superior performance over linear alternatives (such as LDA).