---
title: "Model Training and Evaluation\n"
author: "22110012 - Huu An  \n\n 22110059 - Hong Hien\n"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: false
    theme: united
    highlight: tango
    mathjax: default
  pdata_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include=FALSE}
# --- Global chunk settings ---
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center')

# --- Libraries for modeling and evaluation ---
library(tidyverse)    # Data manipulation and ggplot2
library(caret)        # Machine learning workflow
library(factoextra)   # PCA visualization
library(psych)        # Multivariate statistics
library(biotools)     # Box's M test
library(MVN)          # Multivariate normality testing
library(Rtsne)        # t-SNE
library(uwot)         # UMAP
library(patchwork)    # Arranging plots
library(scales)       # Formatting scales
```

```{r}
# --- Load Extracted Features ---

# Read processed feature set
train_data <- readr::read_csv("EEG_Features_Extracted.csv")

# Separate predictors and target labels
X_features <- train_data %>% dplyr::select(-target)
y_labels_factor <- as.factor(train_data$target)

# Verify dimensions (Expected: 11,500 rows and 39 feature columns)
dim(X_features)
```

In this section, we train, optimize, and benchmark machine learning classifiers on both the full engineered feature space and dimensionality-reduced PCA spaces.

## 6.1. Training Data Preparation

```{r}
# Stratified data partition (80% training, 20% test)
set.seed(42)
trainIndex <- createDataPartition(y_labels_factor, p = 0.8, list = FALSE)

train_features <- X_features[trainIndex, ]
test_features  <- X_features[-trainIndex, ]

y_train <- y_labels_factor[trainIndex]
y_test  <- y_labels_factor[-trainIndex]

# Recode target levels into syntactically valid R variable names for caret
levels(y_train)[levels(y_train) == "0"] <- "NonSeizure"
levels(y_train)[levels(y_train) == "1"] <- "Seizure"
levels(y_test)[levels(y_test) == "0"]   <- "NonSeizure"
levels(y_test)[levels(y_test) == "1"]   <- "Seizure"

# Construct complete training dataframe
train_df <- train_features %>% 
  mutate(y = y_train)
```

```{r}
# Inspect initial training set class distribution
cat("--- Class Distribution in Training Set ---\n")
print(prop.table(table(y_train)))
```

**Observation:**
The Non-Seizure class accounts for ~80% of samples, whereas the Seizure class represents ~20%. This pronounced **class imbalance** risks biasing models toward the majority class, impairing detection of actual seizure episodes—the primary clinical priority.

To address this, we apply **SMOTE (Synthetic Minority Over-sampling Technique)** within our cross-validation recipes to synthetically augment the minority seizure class. This ensures balanced learning dynamics while preventing information leakage into the validation and test splits.

---

## 6.2. Training with 10-Fold Cross-Validation

```{r}
# Configure 10-fold cross-validation control
train_control <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,                 # Required for ROC-AUC computation
  summaryFunction = twoClassSummary, # Reports ROC, Sensitivity, Specificity
  verboseIter = TRUE                 # Display training progress
)
```

```{r}
# Define model suite
models_to_compare <- list()
model_list <- list(LDA = "lda", SVM = "svmRadial", RF = "rf")

# Recipe pipeline for full feature set
full_recipe <- recipe(y ~ ., data = train_df) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors()) %>%
  step_smote(y, over_ratio = 1, seed = 123)

# Train each model under cross-validation
for (model_name in names(model_list)) {
  set.seed(42)
  models_to_compare[[model_name]] <- train(
    full_recipe, 
    data = train_df,
    method = model_list[[model_name]],
    trControl = train_control, 
    metric = "ROC", 
    tuneLength = 3
  )
}

# Aggregate cross-validation resamples
algorithm_comparison <- resamples(models_to_compare)
summary(algorithm_comparison)
```

**Performance Analysis and Algorithm Comparison**

Based on 10-fold cross-validation across LDA, SVM-RBF, and Random Forest (RF):

| Model | ROC (Mean) | Sensitivity | Specificity |
| :--- | :---: | :---: | :---: |
| LDA | 0.9923 | 0.9792 | 0.9359 |
| **SVM** | **0.9980** | **0.9883** | 0.9679 |
| RF | 0.9976 | 0.9845 | **0.9723** |

**Key Insights:**

* **SVM-RBF** achieves the highest **ROC (0.9980)** and **Sensitivity (0.9883)**, demonstrating superior capability in capturing true positive seizure events and minimizing false negatives.
* **Random Forest** attains the highest **Specificity (0.9723)**, showing slightly superior negative class discrimination, with ROC and Sensitivity closely trailing SVM.
* **LDA** exhibits lower performance across all metrics, reflecting the non-linear boundaries in the EEG feature space.

---

```{r}
# Density plot of cross-validated ROC distributions
densityplot(algorithm_comparison, metric = "ROC", auto.key = TRUE,
            main = "Cross-Validated ROC Distribution Across Algorithms")
```

**Observation:**
* **SVM (Green):** Displays a tight, high peak centered around 0.998–0.999, indicating superior accuracy and stability across folds.
* **RF (Yellow):** Highly concentrated distribution, slightly shifted leftward relative to SVM.
* **LDA (Blue):** Broader distribution with lower median performance.

---

```{r}
# Dot plot with 95% confidence intervals
dotplot(algorithm_comparison, metric = "ROC", 
        main = "Average Cross-Validated ROC Comparison (95% CI)")
```

**Observation:**
* **SVM** displays the highest mean ROC with tight, non-overlapping confidence intervals compared to LDA, and statistically significant separation from RF.
* Given these results, **SVM-RBF** is selected as the primary architecture for downstream PCA dimensionality reduction experiments.

## 6.3. Training on PCA-Reduced Feature Spaces

We evaluate the trade-off between model simplicity and classification accuracy by training SVM models across varying numbers of principal components (12, 9, and 4 PCs), comparing them against the full feature model (`SVM_Full`).

```{r}
# Extract baseline full SVM model
svm_full_model <- models_to_compare[["SVM"]]

# Initialize comparison list
pca_comparison_list <- list(SVM_Full = svm_full_model)

# Train SVM across PCA component configurations
for (n_components in c(12, 9, 4)) {
  cat(paste("\n--- Training SVM with", n_components, "Principal Components... ---\n"))
  
  pca_recipe <- recipe(y ~ ., data = train_df) %>%
    step_center(all_predictors()) %>%
    step_scale(all_predictors()) %>%
    step_pca(all_predictors(), num_comp = n_components) %>%
    step_smote(y, over_ratio = 1, seed = 123)
  
  set.seed(42)
  pca_comparison_list[[paste0("SVM_", n_components, "PCA")]] <- train(
    pca_recipe, 
    data = train_df,
    method = "svmRadial",
    trControl = train_control,
    metric = "ROC",
    tuneGrid = svm_full_model$bestTune
  )
}

# Resample comparison
svm_pca_comparison <- resamples(pca_comparison_list)
summary(svm_pca_comparison)
```

**Performance Comparison Across PCA Configurations**

| Model | ROC (Mean) | Sensitivity | Specificity |
| :--- | :---: | :---: | :---: |
| **SVM_Full** | **0.9980** | **0.9883** | **0.9679** |
| SVM_12PCA | 0.9951 | 0.9795 | 0.9614 |
| SVM_9PCA | 0.9946 | 0.9750 | 0.9571 |
| SVM_4PCA | 0.9926 | 0.9694 | 0.9560 |

---

**Impact of Dimensionality Reduction:**

* **SVM_Full** delivers the highest overall performance, demonstrating that retaining all engineered features maximizes discriminatory power.
* **SVM_12PCA and SVM_9PCA** retain exceptional performance ($ROC > 0.994$, $Sensitivity > 0.975$) while compressing feature dimensions by over 69%.
* This confirms that PCA offers an attractive trade-off for computationally constrained environments (e.g., real-time wearable EEG monitoring systems).

---

## 6.4. Evaluation on Independent Test Set

We benchmark the top two candidate architectures—**SVM_Full** and **SVM_12PCA**—on the independent held-out test set ($N = 2,300$).

### 6.4.1. Evaluation of SVM_Full

```{r}
# Predictions on test set using SVM_Full
model_full <- pca_comparison_list[["SVM_Full"]]
predictions_full <- predict(model_full, newdata = test_features)

# Generate confusion matrix
cat("--- SVM_Full Performance on Test Set ---\n")
cm_full <- confusionMatrix(data = predictions_full, reference = y_test, positive = "Seizure")
print(cm_full)
```

**SVM_Full Test Confusion Matrix:**

| | Predicted: NonSeizure | Predicted: Seizure |
| :--- | :---: | :---: |
| **Actual: NonSeizure** | 1827 | 13 |
| **Actual: Seizure** | 13 | 447 |

* **Overall Accuracy:** **98.87%** (95% CI: 98.35% – 99.26%).
* **Sensitivity (Seizure Recall):** **97.17%** (minimizes missed seizures).
* **Specificity (Non-Seizure Recall):** **99.29%** (minimizes false alarms).
* **Cohen's Kappa:** **0.9647** (near-perfect agreement).
* **Balanced Accuracy:** **98.23%**.
* **McNemar's Test ($p = 1.000$):** Symmetric misclassifications ($FN = 13, FP = 13$).

---

### 6.4.2. Evaluation of SVM_12PCA

```{r}
# Predictions on test set using SVM_12PCA
model_pca12 <- pca_comparison_list[["SVM_12PCA"]]
predictions_pca12 <- predict(model_pca12, newdata = test_features)

# Generate confusion matrix
cat("\n--- SVM_12PCA Performance on Test Set ---\n")
cm_pca12 <- confusionMatrix(data = predictions_pca12, reference = y_test, positive = "Seizure")
print(cm_pca12)
```

**SVM_12PCA Test Confusion Matrix:**

| | Predicted: NonSeizure | Predicted: Seizure |
| :--- | :---: | :---: |
| **Actual: NonSeizure** | 1812 | 21 |
| **Actual: Seizure** | 28 | 439 |

* **Overall Accuracy:** **97.87%** (95% CI: 97.19% – 98.42%).
* **Sensitivity:** **95.43%**.
* **Specificity:** **98.48%**.
* **Cohen's Kappa:** **0.9338**.
* **Balanced Accuracy:** **96.96%**.
* **McNemar's Test ($p = 0.3914$):** No significant error asymmetry.

---

### 6.4.3. Test Performance Comparison Summary

```{r}
# Extract performance metrics
metrics_summary <- tibble(
  Metric = c("Accuracy", "Sensitivity", "Specificity", "F1"),
  SVM_Full = c(
    cm_full$overall['Accuracy'],
    cm_full$byClass['Sensitivity'],
    cm_full$byClass['Specificity'],
    cm_full$byClass['F1']
  ),
  SVM_12PCA = c(
    cm_pca12$overall['Accuracy'],
    cm_pca12$byClass['Sensitivity'],
    cm_pca12$byClass['Specificity'],
    cm_pca12$byClass['F1']
  )
)

knitr::kable(metrics_summary, digits = 4, caption = "Performance Comparison on Independent Test Set")
```

| Metric | SVM_Full | SVM_12PCA |
| :--- | :---: | :---: |
| **Accuracy** | **0.9887** | 0.9787 |
| **Sensitivity** | **0.9717** | 0.9543 |
| **Specificity** | **0.9929** | 0.9848 |
| **F1-Score** | **0.9717** | 0.9471 |

---

# VII. Conclusion and Future Directions

This study demonstrates that automated epileptic seizure detection from short-term (1-second) EEG epochs can achieve high diagnostic precision and reliability, establishing a computational foundation to augment traditional visual inspection.

The core of our pipeline is a multi-domain feature engineering framework integrating time domain statistics, frequency domain power spectral densities, and wavelet decomposition coefficients. Multivariate analysis (MANOVA, $p < 2.2 \times 10^{-16}$) and non-linear manifold embeddings (t-SNE, UMAP) rigorously confirm the discriminative validity of the extracted feature space.

To manage computational complexity, Principal Component Analysis was integrated as an efficient reduction strategy. While the reduced model (**SVM_12PCA**) incurs a minor accuracy degradation, it maintains strong performance (**97.87% Accuracy**, **95.43% Sensitivity**), validating PCA as a viable design choice for resource-constrained embedded or wearable architectures.

Among all evaluated paradigms, the **Support Vector Machine with Radial Basis Function kernel on the full feature space (SVM_Full)** demonstrated optimal performance, attaining **98.87% Accuracy**, **97.17% Sensitivity**, and **99.29% Specificity** on the independent test set. This performance profile balances rigorous clinical sensitivity with a minimal false alarm rate.

**Future Research Directions:**
1. Evaluating generalization across multi-center, continuous clinical EEG registries.
2. Developing real-time streaming implementations on ultra-low-power edge hardware.
3. Integrating hybrid deep neural architectures with attention mechanisms to directly ingest raw multi-channel telemetry.
